Kernel Phase and Coronagraphy with Automatic Differentiation

### 2020-07-29 ms.tex

In Figure~\ref{kernel_jacobian} we display the results of our Jacobian calculation for a set of highlighted baselines, showing side by side on the left a baseline superimposed on the support of the optical transfer function (OTF)\footnote{The OTF is the autocorrelation of the pupil, and is in this case also a circle. Its support is calculated by calculating a $u,v$ forwards model with zero phase offset and making an amplitude-threshold mask.}, and on the right next to it a map of the corresponding Jacobian vector (arranged in 2D). This is the transfer matrix $\mathbf{A}_\phi$ arranged row by row.

@@

We calculate the singular values of this matrix (restricted to the support of the pupil) by SVD (Figure~\ref{fig:svd}), finding a sharp cutoff as expected, separating a subspace of kernel phases from phases contaminated by noise from the aberrations. In the nonsingular space, a selection of pairs of corresponding modes in the pupil and $u,v$ plane are displayed in Figure~\ref{nonsingular_modes}, and a selection of kernel phase modes in the $u,v$ plane are shown in Figure~\ref{kernel_modes}. These illustrate an important point: at the edges of the pupil/long baselines there is extra noise. Because of the limited field of view, noise from outside the pupil or OTF is convolved inside, and large elements in the transfer matrix can dominate parts of the SVD. For numerical stability we have therefore excluded the very longest baselines in calculating kernel phases, but it would be preferable to use a more robust SVD that is tolerant of outliers. This effect is not purely a problem with these simulations: this windowing is a common feature of real data: we need to account for the introduction of noise from outside the OTF to correctly calculate kernel phases, as the normal formalism is exact only in the limit of an infinite field of view.

\begin{figure}
\plotone{jacobian.pdf}
\caption{A representation of the Jacobian determined by \textsc{jax} for monochromatic 2$\mu$m diffraction from a 2.0\,m diameter circular pupil. In each pair of images, on the left is the support of the optical transfer function (OTF), with a red dot highlighting the selected baseline; on the right is a map of the corresponding baseline's derivative with respect to phase over the pupil. We see the same maps as in Figure~1 of \citet{martinache10}. \label{kernel_jacobian}}
\end{figure}

\begin{figure}
    \plotone{kerphi_singular.pdf}
     \caption{Log-scale plot of the singular values of the phase transfer matrix versus index for simple monochromatic diffraction (blue), 20\% bandwidth (green). Singular values are normalized to the first singular value. In both cases, there is a sharp cutoff separating significant singular values from a subset within machine precision of zero. The vectors corresponding to these approximately zero singular values span the null space of the phase transfer operator, and correspond to kernel phases.}
    \label{fig:svd}
\end{figure}

@@

We first build a high resolution model of the PHARO pupil using the same code as \citet{martinache20}, and bin it down to a $64\times64$ pixel grid so that pixel values $\in [0,1]$ indicate the fraction of the telescope aperture filling that binned pixel. We then propagate monochromatic 2.145\,$\mu$m light through this onto a $64\times64$-pixel image plane, and use a matrix Fourier transform to map this onto a $u,v$ plane of the same size. We differentiate this using the forwards mode autodiff as above, obtaining the Jacobian shown in Figure~\ref{pharo_jacobian}, and calculate a kernel phase transfer matrix.

\begin{figure}
\plotone{pharo_jacobian.pdf}
\caption{A representation of the Jacobian determined by \textsc{jax} for the PHARO pupil in $K$-band. In each pair of images, on the left is the support of the optical transfer function (OTF), with a red dot highlighting the selected baseline; on the right is a map of the corresponding baseline's derivative with respect to phase over the pupil. We see the same maps as in Figure~1 of \citet{martinache10}. \label{pharo_jacobian}}
\end{figure}

We then extract kernel phases from both $\alpha$~Oph and a point source calibrator $\epsilon$~Her using a version of \textsc{xara} modified to use the \textsc{morphine} matrix FT. Following \citet{martinache20}, we choose as our observables the median kernel phases across all 100 exposures for each, and the standard error of the mean as our base uncertainty on each. To calibrate the kernel phases, we simply subtract those of the calibrator from $\alpha$~Oph and add their uncertainties in quadrature.

To see the effect of the known optical `ghost' (introduced by an unwanted reflection from the neutral density filter used to observe bright stars), we calculate kernel phase `colinearity' maps as in \citet{martinache20}, shown in Figure~\ref{colinearity}. In these maps we can see that the filter ghost shows up strongly as a false binary in the uncalibrated kernel phase maps, but calibrated kernel phases easily remove this and reveal the tight $\alpha$~Oph binary.

@@

We then use Markov Chain Monte Carlo \citep{metropolis53} to sample from the posterior assuming Gaussian likelihoods, using the \texttt{emcee} affine-invariant ensemble sampler \citep{emcee}. We obtain binary parameters of separation $\rho = 115.3 \pm 1.2$~mas, position angle $\theta = 85.6 \pm 0.27$~degrees, and contrast $c = 26.3 \pm 0.6$: these are close to the \citet{martinache20} figures, but differ by much more than their (small) uncertainties. For comparison, the contrast ratio of the system is expected from resolved imaging to be $28.0 \pm 8.3$ in $K$-band \citep{hinkley11}. Posteriors for both inferences are shown in Figure~\ref{comparison_posterior}.

\begin{figure}
\plotone{comparison_posterior.pdf}
\caption{Posterior distributions for the binary parameters of $\alpha$~Ophiuchi as determined by the autodiff pipeline (black) and the \citet{martinache20} implementation (blue). Corner plots produced by \texttt{corner.py} \citep{corner}. \label{comparison_posterior}}
\end{figure}

@@

In order to investigate the effect on both pipelines of wavefront errors leaking in, we generated a cube of 1000 PSFs with 20 Zernike modes of aberrations, this time with 20\,nm standard deviation, for a total peak-to-peak wavefront amplitude of $295\pm 100$\,nm. We then retrieved the system parameters with \texttt{emcee} for the cube, taking as our observables the median kernel phases and the standard error of the mean over the ensemble (posterior in Figure~\ref{comparison_posterior_sim}; and also frame-by-frame using least squares (displayed in Figure~\ref{comparison_posterior_sim_cube}. For a simulated binary with fiducial $\rho = 125$\,mas, $\theta=86^\circ$, and $c = 25$, there is no sign of the large systematic difference between the methods as seen in real data. Both methods, whether frame by frame or on average, retrieve the true parameters at least comparably accurately. In \citet{martinache20} a large error term was added in quadrature in order to account for additional noise in the data unexplained by the diversity over exposures and uncalibrated by the kernel phase model. Whatever systematic this is may be treated differently by these two pipelines, and may need to be a subject for future work.

@@

In numerical experiments, subpixel mismatch in Fourier transform coordinates or even Fourier centering convention could cause mismatches between simulation parameters and retrievals at around the scale of the discrepancy between the two methods.  These results are in line with the \citet{martinache_habilitation} conclusion that inexact Fourier sampling contributes significant systematic noise to kernel phases. The final version used for the autodiff pipeline uses the exact optical model used to generate the data, and a differentiable version of the exact $u,v$ transform implementation used in \textsc{xara}; we do not believe this to be the main cause of the small discrepancy between autodiff and conventional parameters, but it does illustrate that model mis-specification may be a contributing factor.

@@

Calculating the SVD of this Jacobian (again restricted to the support of the primary pupil) we see a decline to very small singular values, albeit not as steep as earlier for the simple imaging case. The ordered singular value curve is displayed in Figure~\ref{fig:svd_coronagraph}.

@@

\begin{figure}
\plotone{kernel_modes_corona.pdf}
\caption{Some kernel speckle maps - damped (nearly-singular) vectors in the image plane from the SVD of the coronagraph Jacobian in~\ref{speckle_jacobian}. There is little apparent structure in these except for weak patterns around the occulted region.
\label{kernel_corona}}
\end{figure}

@@

\begin{figure}
\plotone{speckle_jacobian_zernike.pdf}
\caption{A representation of the Jacobian determined by \textsc{jax} for a simple Lyot coronagraph using a Zernike basis of 200 modes to represent wavefront error. In each pair of images, on the left is the speckle pattern, with a red dot highlighting the selected pixel; on the right is a map of the corresponding pixel's derivative with respect to phase over the pupil. We expect that speckles far from the central dark hole will correspond to sinusoids across the pupil, but in the Zernike basis these appear significantly distorted. \label{speckle_jacobian_zernike}}
\end{figure}

@@

We have re-run the simple diffraction simulation from Section~\ref{sec:simple} with 200 Zernike modes on a $256\times256$ grid, and the coronagraph from Section~\ref{sec:coronagraph} on a $128\times128$ grid using 300 Zernike modes.

The Jacobian derived from this basis for simple diffraction is an excellent approximation to the full calculation on a grid. Given that in problems such as HST data analysis we believe that PSF variations are dominated by low-order modes \citep{pope13}, this allows us to calculate a larger kernel space with a more accurate pupil model than has previously been possible.

On the other hand for the coronagraph, while it is a huge advantage in simulation speed and fidelity, we see in Figure~\ref{speckle_jacobian_zernike} that this basis does not accurately represent the high-spatial-frequency sine waves corresponding to speckles far from the core of the PSF. On the other hand, it does an adequate job representing the effects of low-order wavefront error.

@@

\section{Conclusions}
\label{sec:conclusions}

In this paper we have treated mainly the monochromatic case. It is nevertheless straightforward to include broadband effects as a linear combination of diffraction simulations, which are also differentiable. Likewise, linear mixing of polarization states via Mueller matrices is easily adapted in this framework. This may be important for high angular resolution differential-polarimetry instruments such as SPHERE/ZIMPOL \citep{zimpol} on the VLT, VAMPIRES on the Subaru SCExAO \citep{vampires}, or the GPI polarimeter \citep{gpipol}, where polarization measurements are intrinsic to the goal of the experiment, or to issues with polarization leakage affecting closure phases in holographic aperture masking experiments \citep[e.g.][]{doelman18}.

@@

The approach of \citet{sitzmann2018} to end-to-end optimization of an optical system has clear relevance to optical astronomy. Going beyond using gradients for phase retrieval \citep{jurling14}, we can envision gradient-based \textit{phase design} for coronagraphs and for diffractive-pupil telescopes. For example, it may be useful in rapidly finding deformable mirror settings to generate a ``dark hole'' in high-contrast imaging \citep{malbet95}. It may also be useful for diffractive-pupil design: the \textsc{Toliman} space telescope concept \citep{tuthill18,bendek18} aims to use a pupil-plane phase mask to achieve high precision relative astrometry of $\alpha$~Centauri AB to search for exoplanets. Work so far in designing the phase mask has been in optimizing parametrized pupils with respect to the radially-weighted gradient energy of the PSF as a proxy for astrometric precision. It will be in principle possible to calculate the Fisher information for the astrometric precision, and to use this as an objective function for optimizing a non-parametric \textsc{Toliman} pupil. By differentiating with respect to phase in intermediate planes, \textsc{morphine}-like methods will also be useful in intermediate-plane wavefront control and mask design.

>>>>

In Figure~\ref{kernel_jacobian} we display the results of our Jacobian calculation for a set of highlighted pupil samples and how they influence the phases in the $u,v$ plane.

@@

We calculate the singular values of this matrix (restricted to the support of the pupil) by SVD (Figure~\ref{fig:svd}), finding a sharp cutoff as expected, separating a subspace of kernel phases from phases contaminated by noise from the aberrations. In the nonsingular space, a selection of pairs of corresponding modes in the pupil and $u,v$ plane are displayed in Figure~\ref{nonsingular_modes}, and a selection of kernel phase modes in the $u,v$ plane are shown in Figure~\ref{kernel_modes}. These illustrate an important point: at the edges of the pupil/long baselines there is extra noise. Because of the limited field of view, noise from outside the pupil or optical transfer function (OTF) is convolved inside, and large elements in the transfer matrix can dominate parts of the SVD. For numerical stability we have therefore excluded the very longest baselines in calculating kernel phases, but it would be preferable to use a more robust SVD that is tolerant of outliers. This effect is not purely a problem with these simulations: this windowing is a common feature of real data: we need to account for the introduction of noise from outside the OTF to correctly calculate kernel phases, as the normal formalism is exact only in the limit of an infinite field of view.

% \footnote{The OTF is the autocorrelation of the pupil, and is in this case also a circle. Its support is calculated by calculating a $u,v$ forwards model with zero phase offset and making an amplitude-threshold mask.}
\begin{figure}
\plotone{jacobian.pdf}
\caption{A representation of the Jacobian determined by \textsc{jax} for monochromatic 2$\mu$m diffraction from a 2.0\,m diameter circular pupil. In each pair of images, on the left is the pupil, with a red dot highlighting a selected pixel; on the right is a map of the derivative in the $u,v$ plane with respect to the phase at this point in the pupil. We see the same maps as in Figure~1 of \citet{martinache10}. \label{kernel_jacobian}}
\end{figure}

\begin{figure}
    \plotone{kerphi_singular.pdf}
     \caption{Log-scale plot of the singular values of the phase transfer matrix plotted against index ordered by decreasing singular value for simple monochromatic diffraction (blue), 20\% bandwidth (green). Singular values are normalized to the first singular value. In both cases, there is a sharp cutoff separating significant singular values from a subset within machine precision of zero. The vectors corresponding to these approximately zero singular values span the null space of the phase transfer operator, and correspond to kernel phases.}
    \label{fig:svd}
\end{figure}

@@

We first build a high resolution model of the PHARO pupil using the same code as \citet{martinache20}, and bin it down to a $64\times64$ pixel grid so that pixel values $\in [0,1]$ indicate the fraction of the telescope aperture filling that binned pixel. We then propagate monochromatic 2.145\,$\mu$m light through this onto a $128\times128$-pixel image plane, and use a matrix Fourier transform to map this onto a $u,v$ plane of the same size. We differentiate this using the forwards mode autodiff as above, obtaining the Jacobian shown in Figure~\ref{pharo_jacobian}, and calculate a kernel phase transfer matrix.

\begin{figure}
\plotone{pharo_jacobian.pdf}
\caption{A representation of the Jacobian determined by \textsc{jax} for the PHARO pupil in $K$-band. In each pair of images, on the left is the PHARO pupil, with a red dot highlighting a pupil sample; on the right is a map of the derivative in the $u,v$ plane with respect to phase at that pupil sample. We see the same maps as in Figure~1 of \citet{martinache10}. \label{pharo_jacobian}}
\end{figure}

We then extract kernel phases from both $\alpha$~Oph and a point source calibrator $\epsilon$~Her using a version of \textsc{xara} modified to use the \textsc{morphine} matrix FT. Following \citet{martinache20}, we choose as our observables the median kernel phases across all 100 exposures for each, and the standard error of the mean as our base uncertainty on each. To calibrate the kernel phases, we simply subtract those of the calibrator from $\alpha$~Oph and add their uncertainties in quadrature. To see the effect of the known optical `ghost' (introduced by an unwanted reflection from the neutral density filter used to observe bright stars), we calculate kernel phase `colinearity' maps as in \citet{martinache20}, shown in Figure~\ref{colinearity}. In these maps we can see that the filter ghost shows up strongly as a false binary in the uncalibrated kernel phase maps, but calibrated kernel phases easily remove this and reveal the tight $\alpha$~Oph binary.

In \citet{martinache20} a large error term was added in quadrature in order to account for additional noise in the data unexplained by the diversity over exposures and uncalibrated by the kernel phase model. Using nonlinear least squares to fit the calibrated data, we also find that the best-fitting binary model has a high $\chi^2$. We therefore follow \citet{martinache20} and determine a value for error added in quadrature to make the reduced $\chi^2 = 1.0$. \textcolor{red}{What value?}

@@

We then use Markov Chain Monte Carlo \citep{metropolis53} to sample from the posteriors assuming Gaussian likelihoods, using the \texttt{emcee} affine-invariant ensemble sampler \citep{emcee}. We obtain binary parameters of separation $\rho = 120.2 \pm 1.2$~mas, position angle $\theta = 86.2 \pm 0.3$~degrees, and contrast $c = 24.4 pm 0.6$: all well within 1$\sigma$ of the \citet{martinache20} figures. We likewise use \texttt{emcee} to sample from the likelihood using the \citet{martinache20} code. Posteriors for both inferences are shown in Figure~\ref{comparison_posterior}.

\begin{figure}
\plotone{comparison_posterior.pdf}
\caption{Posterior distributions for the binary parameters of $\alpha$~Ophiuchi as determined by the autodiff pipeline (black) and the \citet{martinache20} implementation (blue). \label{comparison_posterior}}
\end{figure}

@@

In order to investigate the effect on both pipelines of wavefront errors leaking in, we generated a cube of 1000 PSFs with 20 Zernike modes of aberrations, this time with 20\,nm standard deviation, for a total peak-to-peak wavefront amplitude of $295\pm 100$\,nm. We then retrieved the system parameters with \texttt{emcee} for the cube, taking as our observables the median kernel phases and the standard error of the mean over the ensemble (posterior in Figure~\ref{comparison_posterior_sim}; and also frame-by-frame using least squares (displayed in Figure~\ref{comparison_posterior_sim_cube}. For a simulated binary with fiducial $\rho = 125$\,mas, $\theta=86^\circ$, and $c = 25$, there is no sign of the large systematic difference between the methods as seen in real data. Both methods, whether frame by frame or on average, retrieve the true parameters with similar accuracy and precision. %Whatever systematic this is may be treated differently by these two pipelines, and may need to be a subject for future work.

@@

In order to establish the accuracy of both pipelines, we conducted an injection test. We took the first frame of the $\eps$~Her datacube as our `data' and use a subpixel Fourier shift to create a mock binary at a position angle of $90^\circ$, contrast 25, and a range of separations. We then used least squares to fit these data (initialized at the true values), determined in each case an error to add in quadrature to make reduced $\chi^2=1$, and re-ran the least squares fit with new data, saving the best-fit parameters and uncertainties taken from the square root of the diagonal elements of the inverse covariance matrix.

% In numerical experiments, subpixel mismatch in Fourier transform coordinates or even Fourier centering convention could cause mismatches between simulation parameters and retrievals at around the scale of the discrepancy between the two methods.  These results are in line with the \citet{martinache_habilitation} conclusion that inexact Fourier sampling contributes significant systematic noise to kernel phases. The final version used for the autodiff pipeline uses the exact optical model used to generate the data, and a differentiable version of the exact $u,v$ transform implementation used in \textsc{xara}; we do not believe this to be the main cause of the small discrepancy between autodiff and conventional parameters, but it does illustrate that model mis-specification may be a contributing factor.

@@

Calculating the SVD of this Jacobian (again restricted to the support of the primary pupil) we see a decline to very small singular values, albeit not as steep as earlier for the simple imaging case. The ordered singular value curve is displayed in Figure~\ref{fig:svd_coronagraph}. The nonsingular modes form a basis, like the Zernike basis, that maps between aberration patterns in the pupil and the speckle modes they generate. Some pairs of nonsingular modes in the pupil and image plane are shown in Figure~\ref{nonsingular_corona}

@@

% \begin{figure}
% \plotone{kernel_modes_corona.pdf}
% \caption{Some kernel speckle maps - damped (nearly-singular) vectors in the image plane from the SVD of the coronagraph Jacobian in~\ref{speckle_jacobian}. There is little apparent structure in these except for weak patterns around the occulted region.
% \label{kernel_corona}}
% \end{figure}

@@

% \begin{figure}
% \plotone{speckle_jacobian_zernike.pdf}
% \caption{A representation of the Jacobian determined by \textsc{jax} for a simple Lyot coronagraph using a Zernike basis of 200 modes to represent wavefront error. In each pair of images, on the left is the speckle pattern, with a red dot highlighting the selected pixel; on the right is a map of the corresponding pixel's derivative with respect to phase over the pupil. We expect that speckles far from the central dark hole will correspond to sinusoids across the pupil, but in the Zernike basis these appear significantly distorted. \label{speckle_jacobian_zernike}}
% \end{figure}

@@

We have re-run the simple diffraction simulation from Section~\ref{sec:simple} with 200 Zernike modes on a $256\times256$ grid, and the coronagraph from Section~\ref{sec:coronagraph} on a $128\times128$ grid using 300 Zernike modes. 
The Jacobian derived from this basis for simple diffraction is an excellent approximation to the full calculation on a grid. Given that in problems such as HST data analysis we believe that PSF variations are dominated by low-order modes \citep{pope13}, this allows us to calculate a larger kernel space with a more accurate pupil model than has previously been possible.

While it is a huge advantage in simulation speed and fidelity, this basis does not accurately represent the high-spatial-frequency sine waves corresponding to speckles far from the core of the PSF. Nevertheless, it does an adequate job representing the effects of low-order wavefront error, and this may be sufficient for many purposes, for example in optimization of optical design.

@@

\section{Conclusions and Future Work}
\label{sec:conclusions}

We have shown that automatic differentiation software offers significant benefits to astronomical optics. Using a simple model of diffraction, we can recover the idea of kernel phases, and using real Palomar data we obtain similar astrometric performance to the analytically-derived standard approach. Crucially, this can then be extended to arbitrary systems for which analytic self-calibration may not be possible. We have demonstrated this for a simple Lyot coronagraph, finding a modal basis and kernel observables analogous to kernel phase. Though this may be of limited practical importance due to the strong nonlinearities in the Lyot coronagraph problem, other coronagraphs and imaging systems may derive greater benefit.

In this paper we have shown that autodiff allows us to straightforwardly account for broadband effects as a linear combination of diffraction simulations, which are also differentiable. Likewise, linear mixing of polarization states via Mueller matrices could be included in this framework. This may be important for high angular resolution differential-polarimetry instruments such as SPHERE/ZIMPOL \citep{zimpol} on the VLT, VAMPIRES on the Subaru SCExAO \citep{vampires}, or the GPI polarimeter \citep{gpipol}, where polarization measurements are intrinsic to the goal of the experiment, or to issues with polarization leakage affecting closure phases in holographic aperture masking experiments \citep[e.g.][]{doelman18}.

@@

The approach of \citet{sitzmann2018} to end-to-end optimization of an optical system has clear relevance to optical astronomy. Going beyond using gradients for phase retrieval \citep{jurling14}, we can envision gradient-based \textit{phase design} for coronagraphs,  for diffractive-pupil telescopes, or for starshades. For example, it may be useful in rapidly finding deformable mirror settings to generate a ``dark hole'' in high-contrast imaging \citep{malbet95}. It may also be useful for diffractive-pupil design: the \textsc{Toliman} space telescope concept \citep{tuthill18,bendek18} aims to use a pupil-plane phase mask to achieve high precision relative astrometry of $\alpha$~Centauri AB to search for exoplanets. Work so far in designing the phase mask has been in optimizing parametrized pupils with respect to the radially-weighted gradient energy of the PSF as a proxy for astrometric precision. It will be in principle possible to calculate the Fisher information for the astrometric precision, and to use this as an objective function for optimizing a non-parametric \textsc{Toliman} pupil. By differentiating with respect to phase in intermediate planes, \textsc{morphine}-like methods will also be useful in intermediate-plane wavefront control and mask design.
